---
title: "DPI Final Project"
author: "Yashas Mattur"
date: '2022-07-01'
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Code here:

```{r a}
library(tidyverse)
#energy use data set
Energy_Use = Chicago_Energy_Benchmarking <- read.csv("~/Desktop/DPI 2022/Chicago_Energy_Benchmarking.csv", stringsAsFactors=TRUE)
#green roof data set
Green_Roofs <- read.csv("~/Desktop/DPI 2022/Green_Roofs.csv", stringsAsFactors=TRUE)
```

Why does this matter? Background on pollution, urban heat islands, green roofs
- clean data
- graphs (color, 3d, animation, gradient, kde, normal)
  What do these graphs show?
  - bar graph
  - histogram
  - scatter plot
  - regression analysis
    - linear regression
    - logistic regression
- confidence intervals
- hypothesis testing
  - significance test for means (Z/t)
  - significance test for proportions
  - chi-square
    - test for homogeneity
    - test for independence

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

```{r b}
for (x in 2014:2020)
  hist((Energy_Use %>% filter(Data.Year == x))$ENERGY.STAR.Score, xlab = "Energy Star Score", main = x)
```

```{r c}
EnergyPerYear = list()
for (i in 1:7)
  EnergyPerYear[[i]] = Energy_Use %>% filter(Data.Year == i + 2013)
EnergyPerYear[[1]]

```
```{r d}
hist(Energy_Use$Electricity.Use..kBtu.)
```

```{r e}
library(maps)
library(ggplot2)
library(raster)
library(dplyr)
states = c("Illinois", "Indiana", "Wisconsin")
us = getData("GADM", country = "USA", level = 1)
us.states = map_data("state", region = states)
buildings = data.frame(Energy_Use$ENERGY.STAR.Score, Energy_Use$Latitude, Energy_Use$Longitude, Energy_Use$Chicago.Energy.Rating)
base = ggplot() + geom_polygon(data = us.states, aes(x=long, y=lat, group=group), color = "light green", fill = "light green") + geom_path(data = us.states, aes(x=long, y=lat, group=group)) + geom_point(data = buildings, aes(x=Energy_Use.Longitude, y=Energy_Use.Latitude, color=Energy_Use.ENERGY.STAR.Score), size = 1, alpha =I(0.03)) + coord_map() + coord_cartesian(xlim = c(-88.5,-87.5), ylim = c(41.5, 42.5)) + labs(col = "Energy Rating")
cleanup = theme(panel.background = element_rect(fill = "light blue", colour = "white"), axis.line = element_line(colour = "white"), axis.ticks = element_blank(), axis.text.x = element_blank(), axis.text.y = element_blank(), axis.title.x = element_blank(), axis.title.y = element_blank())
base+cleanup
```
```{r f}
hist(Energy_Use$ENERGY.STAR.Score, xlab = "Energy Efficiency", main = "Histogram of Energy Efficiency in Chicago")
```

```{r g}
library(ggplot2)
library(gganimate)
library(gifski)
library(tidyr)
library(sf)
library(transformr)
x = Energy_Use$Electricity.Use..kBtu.
y = Energy_Use$Total.GHG.Emissions..Metric.Tons.CO2e.
s = Energy_Use$District.Chilled.Water.Use..kBtu.
t = Energy_Use$Data.Year
unanimated_plot = ggplot(Energy_Use, aes(x, y, size = s, color = "salmon")) + geom_point(alpha = 0.7, show.legend = F) + stat_smooth(method = "lm", formula = y~x, geom="smooth") + coord_cartesian(xlim = c(0, 250000000), ylim = c(0, 50000)) + scale_size(range = c(2,12)) + labs(title = "Electricity Use vs Emissions in Year: {frame_time}", x = "Electricity Usage", y = "Amount of CO2 Emissions") + transition_time(t) + ease_aes() + theme(legend.position = "none")

gganimate::animate(unanimated_plot, nframes = 7, fps = 2, renderer = gifski_renderer())
anim_save("test.gif")
```

```{r h}
library(patchwork)
hist = ggplot(Energy_Use, aes(x=Total.GHG.Emissions..Metric.Tons.CO2e., y=..density..)) + geom_histogram(binwidth = 1000, color = "blue", fill = "light blue") + transition_states(Data.Year) + labs(title = "Year: {closest_state}", x = "CO2 Emissions (in Metric Tons)", y = "Proportion Density") + geom_density(color = "red") + coord_cartesian(xlim = c(0, 50000))
gganimate::animate(hist)
```

Efficiency is increasing BECAUSE CO2 emissions are decreasing!



```{r i}
#confidence interval for mean of co2 emissions
data = na.omit(Energy_Use$Total.GHG.Emissions..Metric.Tons.CO2e.)
xbar = mean(data)
n = length(data)
c = (1-0.95)/2
sigma = sd(data)*sqrt((n-1)/n)
Z = qnorm(c, lower.tail = F)
xbar + c(-1, 1)*Z*sigma/sqrt(n)
```
```{r}
emissions = na.omit(Energy_Use$Total.GHG.Emissions..Metric.Tons.CO2e.)
emissions = lapply(emissions, \(x) round(x/10000, 0))
sample.emissions = unlist(sample(emissions, 100))
electricity = na.omit(Energy_Use$Electricity.Use..kBtu.)
electricity = lapply(electricity, \(x) round(x/1000000, 0))
sample.electricity = unlist(sample(electricity, 100))
chisq.test(sample.emissions, sample.electricity)
```

```{r j}
#confidence interval for mean of energy score
data = na.omit(Energy_Use$ENERGY.STAR.Score)
xbar = mean(data)
n = length(data)
c = (1-0.95)/2
sigma = sd(data)*sqrt((n-1)/n)
Z = qnorm(c, lower.tail = F)
xbar + c(-1, 1)*Z*sigma/sqrt(n)
```

```{r k}
# h0 - mu = 2014's average Energy Star Score, h1 = mu > 2014's average energy star score
m0 = mean(na.omit((Energy_Use %>% filter(Data.Year == 2014))$ENERGY.STAR.Score))
data = na.omit((Energy_Use %>% filter(Data.Year == 2020))$ENERGY.STAR.Score)
xbar = mean(data)
n = length(data)
s = sd(data)
df = n-1
tts = (xbar - m0)/(s / sqrt(n))
pt(tts,df,lower.tail = TRUE)
# reject h0, meaning that the average energy star score has risen since 2014. 
```


